---
title: "Day 3 Homework"
output: html_document
---

1. Identify a column in your data that *should* be a factor (whether it is or not). Check to see whether R is already treating it as a factor, and if not, convert it into a factor.

```{r}

durdisc.group <- durdisc.data$Experimental.Group
is.factor(durdisc.group)
durdisc.group.factor <- as.factor(durdisc.group)

```

2. What are the levels of the factor? How many occurrences of each level are there in the data?
   
```{r}

levels(durdisc.group.factor)
table(durdisc.group.factor) # builds a contingency table

```

3. Create a new column that is a numeric "version" of the factor column, using `as.numeric()`. Use `xtabs()` to see how the values of these columns line up.

```{r}

durdisc.group.numeric <- as.numeric(durdisc.group.factor)
xtabs(durdisc.group.numeric ~ durdisc.group.factor) # im getting an answer but it seems wrong

```

4. Create another column that is a copy of the factor column. Now, recode this new column so that the factor levels have new names. 
  (e.g., if you have a factor with levels "A" and "B", you could make a new column that had "condition A" and "condition B" as the level names instead)

```{r}

durdisc.group.factor2 < - levels(durdisc.group.factor)[1] <- "ActiveCtrl"
durdisc.group.factor2 < - levels(durdisc.group.factor)[2] <- "ExpTraining"

##this didn't seem to work. where was i supposed to add the column? to my data.frame?

```

5. Change the order of the levels in your new factor, using three different methods:
  a. `relevel()`
  b. `reorder()`
  c. `factor()`

```{r}

#using my old factor because the prior one didn't work

relevel(durdisc.group.factor, ref = "Training")
with(durdisc.group.factor, reorder(??)) ## i can't figure out what to put in here
as.ordered(durdisc.group.factor) # ack not working either

```


