---
title: "Day 4 Lecture"
output: html_document
---

# Today's agenda

  - Review & questions
  - Character/string data
  - Factors
  - Excursion: simple data frame manipulation, `xtabs()`
  - PRACTICE: working with factors, part 1
  - Special values: NA/NULL/Inf/NaN
  - PRACTICE: data type coersion
  - Lists
  - Creating data frames from scratch
  - Combining data frames
  - PRACTICE: binding & merging

# Review
  - Go over Day 3 Practice
  - Day 3 Homework example
  - Knitting notes

# Characters
  - more primitive than factors
  - can manipulate with regular expressions, other text functions

Some handy text stuff

```{r}

#quotes

x <- "a string"
y <- "also a string"
x
y

# cat(), print()

print(x)
cat(x)
cat(x, file = "cat_example.txt")

sample.ttest <- t.test(rnorm(20, 0, 1), rnorm(20,2,1))
sample.ttest #not an object, this is a result of printing an object
print(sample.ttest)
printed.ttest <- print(sample.ttest) #saving that actual printed result as an object

# special characters: \t, \n, "escaping"

x<- "here's a cool result:\n\n\t10" # line returns (n) and tabs (t)
cat(x)

cat("some\\ntext") # need to use a double to keep a real slash in your text


# nchar()

nchar("hello") #num of characters
length("hello") #num of elements

# paste(), strsplit()

paste("hello", "there") #pastes different arguments together

x <- c("here", "are")
x
paste("prefix", x)
paste("prefix", x, sep = "")
paste(x, collapse = " ")

# stringr package

# grep(), grepl(), gsub()

x
grep("here", x) #where the string was found in the vector; can do letters too
grep1("here", x)
gsub("e", "E", x)

longer.thing <- c("10F503", "356G650", "20F245")
gsub("\\d+\\w(\\d+)", "\\1", longer.thing)

```

# Factors
  - Actually numbers under the hood
  - Act like numbers + labels
  - VERY useful for many kinds of analyses
  - if a column is a *property* of data (not a measure itself), factors are usually best
    - groupings = good as factors
    - conditions = good as factors
  - if a column is *data*, maybe consider keeping it a string
    - free response answers = bad as factors
  - `as.factor()` coerces to a factor
  - `factor()` creates a factor
  - factors have *levels*: categories
  - `stringsAsFactors` : option for how to read in character vectors
    - = FALSE if you don't want to read in characters as factors

```{r}
# factor()

x<- c(rep("a", 20), rep("b", 20), rep("c", 20))
x   
summary(x)
x.factor <- factor(x)
summary(x.factor)

# levels()

levels(x.factor)


# summary()

# coersion

# subsetting

sleep
mysleep <- sleep
summary(mysleep) #10 cases of 1 and 10 cases of 2 in group; r thinks this is a factor
class(mysleep$group)
levels(mysleep$group)

mysleep.group1 <- mysleep[mysleep$group == "1",]
summary(mysleep.group1) #zeros out group label "2"

# droplevels

mysleep.group1 <- droplevels(mysleep.group1) #gets rid of factor levels with no observations
summary(mysleep.group1)

# changing levels (labels)
mysleep <- sleep
levels(mysleep$group)

levels(mysleep$group) <- c("pre", "post")
levels(mysleep$group)
mysleep
head(mysleep)

colnames(mysleep) <- c("extra sleep", "treatment group", "subject")
head(mysleep)

# re-ordering

mysleep <- sleep
levels(mysleep$group) <- c("pre", "post")
summary(mysleep)

#levels(mysleep$group) <- c("post", "pre") # CHANGES DATA, NOT ORDER OF LEVELS ....this is bad!!!

mysleep$group2 <- mysleep$group #makes a new column 
head(mysleep)
mysleep$group2 <- factor(mysleep$group2, levels = c("post", "pre")) #actually changes levels; can 
#use this if you want to change a reference level
summary(mysleep)

```

# Data manipulation excursion
## Creating (simple) new columns

```{r}
# single value
mysleep$experiment <- 1
mysleep$group3 <- mysleep$group #assigns new column 

# using subsetting
summary(mysleep)

# ifelse()

mysleep$sleeplevel <- ifelse(mysleep$extra >= median(sleep$extra), "high", "low")
summary(mysleep)
mysleep$sleeplevel <- as.factor(ifelse(mysleep$extra >= 1, "high", "low")) #high will come first (bad)
summary(mysleep)
mysleep$sleeplevel <- factor(ifelse(mysleep$extra >= 1, "high", "low"), levels = c("low", "high"))
summary(mysleep)

```

## Using xtabs() to check coding/cross-tabs

```{r}

xtabs(~group +group2, data = mysleep)
xtabs(~sleeplevel + extra, data = mysleep)

```

# PRACTICE: #0 -- #6

# NAs, NULLs

## NAs
  - missing data, whereas NULLs mean it doesn't exist
  - can be a huge pain
  - hugely important to handle properly!
  - no universal "handle NAs" or "ignore NAs" argument
  - sometimes `na.rm = TRUE` (like for `mean()`)
  - sometimes it's other stuff
  - `cor()` use = "pairwise.complete.obs"
  - no universal "deal with NAs" argument
  
```{r}



```

## NULL
  - another special value
  - means "nothing" or "doesn't exist" (contrast with "missing" or "unknown")

```{r}

```

## Inf, -Inf, NaN
  - can sometimes result from mathematical operations
  - helpful to be able to identify, locate them

```{r}

#dividing by zero, etc will return an Inf
#dividing by a negative sq root will return a NaN 

```

# Data types, classes, & coersion

```{r}
# typeof()

# class()

```

`is.X` family of functions (`is.factor()`, `is.na()`, `is.finite()`, etc.)
`as.X` family of functions (`as.factor()`, `as.numeric()`, `as.data.frame()`, etc.)

# PRACTICE: #7

# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

```

# PRACTICE: #8 -- #11

# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}

```

## `merge()` and `*_join()` from `dplyr` package
```{r}

```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details



# PRACTICE: #12 -- #14


